Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2016-08-23 01_introduction/introduction.tex
Review the introduction and data

\Ac{cap} is the second most frequently diagnosed men cancer, accounting for 899,000 cases leading to 258,100 deaths~\citep{ferlay2010estimates}.
As highlighted by the PI-RADS Steering Committee, the two main challenges to be addressed are~\citep{weinreb2016pi}:
(i) the improvement of detecting clinically significant \ac{cap} and
(ii) an increase of the confidence in benign or dormant cases, avoiding unnecessary invasive medical exams.
In this regard, \ac{mpmri} is frequently used to build robust \ac{cad} systems to detect, localize, and grade \ac{cap}.
In general, \ac{cad} systems are based on \ac{mpmri} which combines several of the following modalities~\citep{lemaitre2015computer}: \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri}, \ac{adc} maps, and \ac{mrsi}.

In \ac{dce}-\ac{mri}, a contrast media is injected intravenously and a set of images is acquired over time.
Consequently, each voxel in the image is a dynamic signal which is related to the vascular properties of the tissue.
In fact, these properties are automatically extracted using quantitative or semi-quantitative approaches~\citep{lemaitre2015computer}.

The former group of approaches uses pharmacokinetic modelling based on a bicompartment model, namely Brix~\citep{brix1991pharmacokinetic} and Tofts~\citep{tofts1995quantitative} models.
The parameters of the Brix model are found assuming a linear relationship between the media concentration and \ac{mri} signal intensity.
This assumption has shown, however, to lead to inaccurate parameter calculation~\citep{heilmann2006determination}.
In the contrary, Tofts model only requires a conversion from \ac{mri} signal intensity to concentration, which can become a non-linear relationship using specific equation of \ac{mri} sequences (e.g., FLASH sequence).
Tofts modelling suffers, however, from an higher complexity~\citep{gliozzi2011phenomenological}.
The conversion using the non-linear approach requires to acquire a T$_1$ map which is not always possible during clinical examination.
Furthermore, the parameter calculation require the \ac{aif} which is challenging to measure and can also lead to inaccurate estimation of the parameters.

@@

Nevertheless, all presented methods suffer from two major drawbacks:
(i) the inter-patient variability of the data lead to a variation of the parameters estimated and to poor classification performance while designing \ac{cad} systems, and
(ii) only few parameters are used to characterize the dynamic signal implying that some information are discarded.

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduce the inter-patient variability of the data.
Furthermore, we show that using the full normalized signal lead to the best classification performance.

>>>>

\Ac{cap} is the second most frequently diagnosed men cancer, accounting for 899,000 cases leading to 258,100 deaths~\citep{ferlay2010estimates}.
As highlighted by the PI-RADS Steering Committee, the two main challenges to be addressed are~\citep{weinreb2016pi}:
(i) the improvement of detecting clinically significant \ac{cap} and
(ii) an increase of the confidence in benign or dormant cases, avoiding unnecessary invasive medical exams.
In this regard, \ac{mpmri} is frequently used to build robust \ac{cad} systems to detect, localize, and grade \ac{cap}.
In general, \ac{cad} systems are based on \ac{mpmri} which potentially combines several of the following modalities~\citep{lemaitre2015computer}: \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri}, \ac{adc} maps, and \ac{mrsi}.

In \ac{dce}-\ac{mri}, a contrast media is injected intravenously and a set of images is acquired over time.
Consequently, each voxel in an image corresponds to a dynamic signal is related to the vascular properties of the tissue.
In fact, these properties are automatically extracted using quantitative or semi-quantitative approaches~\citep{lemaitre2015computer}.

The former group of approaches uses pharmacokinetic modelling based on a bicompartment model, namely Brix~\citep{brix1991pharmacokinetic} and Tofts~\citep{tofts1995quantitative} models.
The parameters of the Brix model are inferred assuming a linear relationship between the media concentration and the \ac{mri} signal intensity.
This assumption has shown, however, to lead to inaccurately estimate the pharmacokinetic parameters~\citep{heilmann2006determination}.
In the contrary, Tofts model requires a conversion from \ac{mri} signal intensity to concentration, which become a non-linear relationship using specific equation of \ac{mri} sequences (e.g., FLASH sequence).
Tofts modelling suffers, however, from an higher complexity~\citep{gliozzi2011phenomenological}.
Indeed, the conversion using the non-linear approach requires to acquire a T$_1$ map which is not always possible during clinical examination.
Additionally, the parameter calculation requires the \ac{aif} which is challenging to measure and can also lead to inaccurate estimation.

@@

Nevertheless, all presented methods suffer from two major drawbacks:
(i) the inter-patient variability of the data lead to a variation of the parameters estimated and subsequently to poor classification performance while designing \ac{cad} systems, and
(ii) only few parameters are used to characterize the dynamic signal implying that some information are discarded.

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduce the inter-patient variability of the data.
Furthermore, we show that using the full normalized \ac{dce}-\ac{mri} signal lead to the best classification performance.

### 2016-08-23 02_methods/methods.tex
Review the introduction and data

The multi-parametric \ac{mri} data are acquired from a cohort of patients with higher-than-normal level of \ac{psa}.
The acquisition is performed using a 3T whole body \ac{mri} scanner (Siemens Magnetom Trio TIM, Erlangen, Germany) using sequences to obtain \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri} and \ac{dw}-\ac{mri}.
Aside of the \ac{mri} examination, these patients also have underwent a guided-biopsy.
The dataset is composed of a total of 20 patients of which 18 patients have biopsy proven \ac{cap} and 2 patients are ``healthy'' with negative biopsies.
Therefore, 13 patients have a \ac{cap} in the \ac{pz}, 3 patients have \ac{cap} in the \ac{cg}, 2 patients have invasive \ac{cap} in both \ac{pz} and \ac{cg} and finally 2 patients are considered as ``healthy''.
An experienced radiologist has segmented the prostate organ --- on \ac{t2w}- and \ac{dce}-\ac{mri} --- as well as the prostate zones (i.e., \ac{pz} and \ac{cg}) and \ac{cap} on the \ac{t2w}-\ac{mri}.

The \ac{dce}-\ac{mri} sequence consists in a kinetic study composed of 40 samples over time with a time resolution of \SI{6.5}{\second}.
These \ac{dce}-\ac{mri} sequences are resampled using the spatial information of the \ac{t2w} \ac{mri} sequence with dimensions of $448 \times 360 \times 64$ and voxel spacing of $0.68 \times 0.68 \times 1.25 $ mm\textsuperscript{3}.
A linear interpolation is used to compute missing data during the up-sampling.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered, to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}- and \ac{dce}-\ac{mri} in order to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

In this work, we proposed a method to normalized \ac{dce}-\ac{mri} prostate data, although it can be ported to any \ac{dce}-\ac{mri} sequences.
The aim of the method is to reduce the intra-patient variations that can occur during acquisition.
In \ac{t2w}-\ac{mri} prostate image, an offset and a scale factor are the two factors driving the intra-patient variation.
Therefore, these variations are corrected using a $z$-score approach assuming that the data follow a Rician distribution~\citep{lemaitre2016normalization}.
In \ac{dce}-\ac{mri}, there is additional factors characterizing such variations.
These variations can be highlighted by observing the evolution of the intensity \ac{pdf} of the \ac{dce}-\ac{mri} over time, as shown in the heatmap in Fig.\,\ref{fig:heatmap}.
These variations are:
(i) an offset of the peak value before pre-contrast,
(ii) a time offset depending of the contrast injection, and
(iii) a global scale factor related to the enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

>>>>

The multi-parametric \ac{mri} data are acquired from a cohort of patients with higher-than-normal level of \ac{psa}.
The acquisition is performed using a 3T whole body \ac{mri} scanner (Siemens Magnetom Trio TIM, Erlangen, Germany) using sequences to obtain \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri} and \ac{dw}-\ac{mri}.
Aside of the \ac{mri} examination, these patients also have underwent a guided-biopsy.
The dataset is composed of a total of 20 patients of which 18 patients have biopsy proven \ac{cap} and 2 patients are ``healthy'' with negative biopsies.
Therefore, 13 patients have a \ac{cap} in the \ac{pz}, 3 patients have \ac{cap} in the \ac{cg}, 2 patients have invasive \ac{cap} in both \ac{pz} and \ac{cg} and finally 2 patients are considered as ``healthy''.
An experienced radiologist has segmented the prostate organ --- on \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} --- as well as the prostate zones (i.e., \ac{pz} and \ac{cg}) and \ac{cap} on the \ac{t2w}-\ac{mri}.

A \SI{3}{\mm} slice fat-suppressed \ac{t2w} fast spin-echo sequence (\ac{tr}/\ac{te}/\ac{etl}: \SI{3400}{\ms}/\SI{85}{\ms}/13) is used to acquire images in sagittal and oblique coronal planes, the latter planes being orientated perpendicular or parallel to the prostate \ac{pz} – rectal wall axis.
Three-dimensional \ac{t2w} fast spin-echo (\ac{tr}/\ac{te}/\ac{etl}: \SI{3600}{\ms}/\SI{143}{\ms}/109, slice thickness: \SI{1.25}{\mm}) images are then acquired in an oblique axial plane.
The nominal matrix and \ac{fov} of the 3D \ac{t2w} fast spin-echo images are $320 \times 256$ and $280 \times 240$ mm\textsuperscript{2}, respectively, thereby affording sub-millimetric pixel resolution within the imaging plane.

\ac{dce}-\ac{mri} are performed using a fat suppressed 3D T1 VIBE sequence (\ac{tr}/\ac{te}/Flip angle: \SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times 192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial information of the \ac{t2w}-\ac{mri} and missing data are interpolated using linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered, to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} in order to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

In this work, we propose a method to normalize \ac{dce}-\ac{mri} prostate data, although it can be applied to any \ac{dce}-\ac{mri} sequences.
The aim of the method is to reduce the intra-patient variations, occuring during the acquisition process.
In \ac{t2w}-\ac{mri}, these variations imply that the intensity \ac{pdf} of the prostate region is shifted as well as wider or narrower, from a patient to another.
Therefore, these variations can be corrected using a $z$-score approach, assuming that the data follow a specific distribution~\citep{lemaitre2016normalization}.

In \ac{dce}-\ac{mri}, the inter-patient variations are more complex due to the over time acquisition.
These variations can be highlighted by observing the evolution of the intensity \ac{pdf} of the \ac{dce}-\ac{mri} over time, as shown in the heatmap in Fig.\,\ref{fig:heatmap}.
Therefore, these variations are:
(i) an offset of the peak value before pre-contrast,
(ii) a time offset depending of the contrast injection, and
(iii) a global scale factor related to the enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.
